From Reversible Programs to \\ Univalent Universes and Back

### 2017-06-27 upi.tex
sec 3 in progress

The syntactic category $c$ is that of 1-combinators denoting
reversible programs, type isomorphisms, permutations, or equivalences
depending on one's favorite interpretation. There are two primitive
combinators $\id$ and $\swap$ which are closed under inverses $!$ and
sequential composition $\circ$. The syntactic category $\alpha$ is
that of 2-combinators denoting reversible program transfomations,
coherence conditions on type isomorphisms, equivalences between
permutations, or program optimizations depending on one's favorite
interpretation.

@@

@@

\begin{verbatim}
Thm: Let P : Type -> Type such that for all X : Type, P(X)
is a proposition. Then Σ(X,P) is the base of a univalent
fibration.

@@

is-prop : Type -> Type has the property that is-prop(X)
is a proposition for all types X. So Σ(X , is-prop) is
the base of a univalent fibration. In other words, this
is the subuniverse consisting of all and only propositions.

is-set : Type -> Type has the property that is-set(X)
is a proposition for all types X. So Σ(X , is-set) is
the base of a univalent fibration. In other words, this
is the subuniverse consisting of all and only sets.

is-1-type : Type -> Type has the property that
is-1-type(X) is a proposition for all types X. So
Σ(X , is-1-type) is the base of a univalent fibration.
In other words, this is the subuniverse consisting of all
and only 1-types. (Externally, these are the 1-groupoids
embedded in inf-groupoids.)

>>>>

The syntactic category $c$ is that of 1-combinators denoting
reversible programs, type isomorphisms, permutations, or equivalences
depending on one's favorite interpretation. There are two primitive
combinators $\id$ and $\swap$ which are closed under inverses $!$ and
sequential composition $\circ$. The syntactic category $\alpha$ is
that of 2-combinators denoting reversible program transformations,
coherence conditions on type isomorphisms, equivalences between
permutations, or program optimizations depending on one's favorite
interpretation.

@@

We now switch gear and focus on constructions of \emph{univalent
  universes}. In order to state the definition of such universes we
need to introduce two notions: propositional equality
$\AgdaDatatype{≡}$ and equivalence $\AgdaDatatype{≃}$.

Propositional equality, or identity types in intensional dependent
type theory, or path types under the HoTT-interpretation come equipped
with a single constructor \AgdaInductiveConstructor{refl} and an
induction principle $\AgdaFunction{J}$. A peculiar aspect of
$\AgdaFunction{J}$ is that it does \emph{not} enforce that every
element of an identity type
$\AgdaFunction{x}~\AgdaDatatype{≡}~\AgdaFunction{y}$ is
\AgdaInductiveConstructor{refl}. This ``weakness'' has a beautiful
consequence: instead of identifying all proofs and collapsing the
structure of proofs in the process, it instead permits a rich
combinatorial structure of proofs to emerge. The second notion we need
is that of equivalence. We say that two types or spaces $A$ and $B$
are equivalent if there exist functions $f : A \rightarrow B$ and
$g : B \rightarrow A$ that compose to the identity function. A notable
point is that equivalences do not come with an induction principle. We
say a universe $\mathcal{U}$ is univalent if its space of identities
$\AgdaFunction{A} ≡ \AgdaFunction{B}$ is equivalent to its space of
equivalences $\AgdaFunction{A} ≃ \AgdaFunction{B}$.

Instead of talking about univalent universes, we generalize the
discussion to univalent type families which will allow the
constructions of custom universes.

\newpage

@@

\begin{verbatim}
Thm: Let P : Type -> Type such that for all X : Type, P(X)
is a proposition. Then Sigma(X,P) is the base of a univalent
fibration.

@@

is-prop : Type -> Type has the property that is-prop(X)
is a proposition for all types X. So Sigma(X , is-prop) is
the base of a univalent fibration. In other words, this
is the subuniverse consisting of all and only propositions.

is-set : Type -> Type has the property that is-set(X)
is a proposition for all types X. So Sigma(X , is-set) is
the base of a univalent fibration. In other words, this
is the subuniverse consisting of all and only sets.

is-1-type : Type -> Type has the property that
is-1-type(X) is a proposition for all types X. So
Sigma(X , is-1-type) is the base of a univalent fibration.
In other words, this is the subuniverse consisting of all
and only 1-types. (Externally, these are the 1-groupoids
embedded in inf-groupoids.)
